---
title: "L5: IP Weighting and G-Computation"
author: "Jean Morrison"
institute: "University of Michigan"
date: "Lecture on 2024-02-07 \n(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
$\newcommand{\ci}{\perp\!\!\!\perp}$
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_search(show_icon = TRUE)
xaringanExtra::use_panelset()
```

```{r xaringanthemer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  link_color = "#ea8a1a",
  base_color = "#110566",
 # header_font_google = google_font("Josefin Sans"),
 # text_font_google   = google_font("Montserrat", "300", "300i"),
 # code_font_google   = google_font("Fira Mono")
)
```

## Lecture Plan 

1. Intro

2. Inverse Probability Weighting 
  - IP weighting for binary and continuous treatment
  - Marginal structural models
  - Censoring weights

3. G-Computation

---
# 1. Introduction

---
## Modeling

- So far we have learned two strategies for estimating $E[Y(a)]$ from data when we need to adjust for
confounding or selection bias. 

- In IP weighting we re-weight our observations by $1/P[A = A_i \vert L_i]$, the probability that 
they received the treatment that they got given confounderts $L_i$. 

- In standardization, we first stratify the data by $L$, compute $E[Y \vert A = a, L]$ within each stratum, and then take a weighted sum of these averages. 

---

## Models


- Formally, a **model** is a class of probability distributions $\mathcal{M} = \lbrace P_\theta : \theta \in \Theta \rbrace$. 

- $\theta$ is a (possibly infinite) vector of parameters and $\Theta$ is the parameter space. 

- A model is called **parametric** when $\theta$ is finite dimensional and $\mathcal{M}$ is not the set of all possible probability distributions. 

- If $\theta$ is infinite dimensional, $\mathcal{M}$ may be called **non-parametric** or **semi-parametric**.
  + Semi-parametric models have a finite, structured component, and an infinite component. 
  
- In this  lecture we will talk about parametric models. 

---

## Our Models So Far

- So far in examples, we have always used **saturated** models. 
  + Saturated models contain all possible probability distributions for the observed data. 
  

- We assume that $A$, $Y$, and any confounders have a discrete number of levels and that we have 
enough data to estimate quantities like $E[Y \vert A, L]$ in each stratum of $A$ and $L$. 
  

- This has allowed us to focus on **identification**, i.e. could we estimate a parameter with infinite data. 

- However, in the real world, we don't get infinite data and it is impractical to assume we can always estimate $E[Y \vert A, L]$ in every stratum of $A$ and $L$.

- We will need **modeling** to go further. 


---

## G-Formula

- The formula we saw previously as the "stratification" formula is also called the g-formula

$$
E[Y(a)] = \sum_l E[Y \vert A = a, L = l]P[L = l]
$$

- We showed in L1 that an alternate expression of the g-formula is 

$$
E[Y(a)] = E\left[ \frac{I(A = a) Y}{f(A \vert L)}\right]
$$

- We can use either of these expressions to obtain parametric estiamtes of $E[Y(a)]$. 

- In G-computation (also called standardization) we estimate $E[Y \vert A, L]$ and plug into the first expression. 

- In IP weighting we estimate $f(A \vert L)$ (the propensity score) and use the second expression. 

---
# 2. Inverse Probabilty Weighting 

---
## NHEFS Data

- Cigarette smokers aged 25-74 were recruited around 1971 and given a survey. 

- Ten years later, participants were given a followup survey. 

- We are interested in estimating the effect of quitting smoking on weight change on the additive scale. 


---

## Variables

- Our exposure is binary (whether or not a person quit smoking between the first and second survey).

- The outcome is the change in weight in kg. 


- What other features would you want to know in order to estimate the causal effect?

--

- For now we will assume that the following variables are sufficient: 

+ Sex (0: male, 1: female)
+ Age (in years)
+ Race (0: white, 1: other)
+ Education (5 categories)
+ Intensity of smoking (cigarettes per day)
+ Duration of smoking (years)
+ Physical activity in daily life (3 categories)
+ Recreational exercise (3 categories)
+ Weight (kg)

---
## Limitations

- For now we will focus on individuals with complete data. 
  + There are some people who did not fill in the second survey. 
  + For these people we know covariates but don't know either the exposure or the outcome. 
  + Additionally, there are 63 people who did fill in the second survey but who's weight is unknown. 
  + More on this later. 

- Individuals may have quit at different times. 

  + We could think of $A$ as a time-varying treatment (coming up later). 

---
## Simple Analysis

- The simplest analysis is to simply compare the mean weight change between quitters and non-quitters. 


```{r, echo = FALSE, message = FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
dat.u <- read_csv("../../data/nhefs.csv") 
dat <- dat.u %>%
       filter(!is.na(wt82))
tab <- dat %>% group_by(qsmk) %>% filter(!is.na(wt82_71)) %>% summarize( mean_wt = mean(wt82_71, na.rm=T), n = n())

kable(tab, col.names = c( "Quit Smoking", "Average Weight Change", "N"), 
      format = 'html', row.names = FALSE, digits = 2) 
```

- On average, quitters gained about 5kg over the 10 year period while non-quitters only gained about 2kg. 

---
## Quitters v Non-Quitters

```{r, echo = FALSE}

tab1 <- dat %>% group_by(qsmk) %>%
        summarize(age_m = mean(age), age_sd = sd(age), 
                  fem_n = sum(sex), fem_p = mean(sex)*100, 
                  race_n = sum(race), race_p = mean(race)*100, 
                  ed5_n = sum(education==5), ed5_p = mean(education==5)*100,
                  wt_m = mean(wt71), wt_sd = sd(wt71),
                  smokeint_m = mean(smokeintensity), smokeint_sd = sd(smokeintensity), 
                  smokeyrs_m = mean(smokeyrs), smokeyrs_sd = sd(smokeyrs), 
                  ex2_n = sum(exercise==2), ex2_p = mean(exercise==2)*100, 
                  act2_n = sum(active ==2), act2_p = mean(active==2)*100, 
                  Total = n())

tab1_f <- tab1 %>% 
          mutate(age = paste0(round(age_m, digits = 1), " (", round(age_sd, digits = 1), ")"), 
                 female = paste0(round(fem_p, digits = 1), "% (", fem_n, ")"), 
                 nonwhite = paste0(round(race_p, digits = 1), "% (", race_n, ")"), 
                 university = paste0(round(ed5_p, digits = 1), "% (", ed5_n, ")"), 
                 weight71 = paste0(round(wt_m, digits = 1), " (", round(wt_sd, digits = 1), ")"), 
                 int = paste0(round(smokeint_m, digits = 1), " (", round(smokeint_sd, digits = 1), ")"), 
                 yrs = paste0(round(smokeyrs_m, digits = 1), " (", round(smokeyrs_sd, digits = 1), ")"), 
                 ex2 = paste0(round(ex2_p, digits = 1), "% (", ex2_n, ")"),
                 act2 = paste0(round(act2_p, digits = 1), "% (", act2_n, ")")
                 ) %>%
          select(qsmk, age, female, nonwhite, university, weight71, int, yrs, ex2, act2, Total) %>% 
          gather("variable", "value", -qsmk) %>% spread(qsmk, value) 

o <- match(c("Total", "age", "female", "nonwhite", "int", "yrs", "weight71", "university", "ex2", "act2"), tab1_f$variable)
tab1_f <- tab1_f[o,]
tab1_f$variable <- recode(tab1_f$variable, "act2" = "Inactive life", 
                          "age" = "Age, years",
                          "nonwhite"="Non-White", 
                          "weight71" = "Weight, kg", 
                          "female" = "Female", 
                          "int" = "Cigarettes/day", 
                          "university" = "College",
                           "yrs" = "Years smoking", 
                          "ex2" = "Little exercise")
kable(tab1_f, col.names = c("Variable", "Did Not Quit Smoking", "Quit Smoking"), 
      format = 'html', row.names = FALSE, digits = 2) 
```


---
## Estimating Weights

- In our data, $L$ is a vector of nine measurements. 

- We cannot compute $P[A = 1 \vert L ]$ among every (or any) stratum of $L$ because every
participant has their own unique value of $L$. 

- We have to fit a parametric model. 

- What would you do? 

---

## Weights Estimated by Logistic Regression 

- We will use a logistic regression model to predict $P[A \vert L]$

- The goal is to get the best possible estimate of the probability of treatment given $L$, so 
we should fit a flexible model. 

- Once we fit the model, we can estimate the probability of quitting for each person as  $\pi_i = P[A = 1 \vert L = L_i]$. These are the probability-scale fitted values from the logistic model. 

- The weight for individual $i$ will be equal to $\frac{1}{\pi_i}$ if $A_i = 1$ (the person actually did quit) or $\frac{1}{1-\pi_i}$ if $A_i = 0$ (the person did not quit). 

---

## Weights Estimated by Logistic Regression 

```{r}
fit <- glm(qsmk ~ sex + race + age + I(age ^ 2) +
    as.factor(education) + smokeintensity +
    I(smokeintensity ^ 2) + smokeyrs + I(smokeyrs ^ 2) +
    as.factor(exercise) + as.factor(active) + wt71 + I(wt71 ^ 2),
    family = binomial(), data = dat)

dat$w <- ifelse(dat$qsmk == 0, 
                1/(1 - predict(fit, type = "response")), 
                1/(predict(fit, type = "response")))
```

```{r, echo = FALSE}
summary(dat$w)
```


- Why is the mean value close to 2?

---

## Positivity 

- *Structural violations* of positivity occur when it is impossible for people with some 
levels of confounders to receive a particular level of treatment. 

  + If we have structural violations, we cannot use IP weighting or standardization. 
  + We need to restrict our inference to relevant strata of $L$. 

- *Random violations* of positivity occur when certain combinations of $L$ and $A$ 
are missing from our data by chance. 

  + We are using a parametric model, so we are able to smooth over unobserved covariate values. 
  + We are able to predict for strata that were not observed in our data. 
  + We should be careful about predicting outside the range of the observed data. 


---

## Assessing Positivity in Propensity Scores

- If positivity holds, propensity scores should be bounded away from 0 and 1:
  
  + Scores very close to 0 or very close to 1 suggest that there are some strata of $A$ and $L$ that have no chance to receive one of the two treatments.
  + We might get scores close to 0 or 1 if there is perfect separation by one or a combination of confounders.
  + Scores close to 0 or 1 suggest there could be structural positivity violations. 

- Propensity scores should have approximately the same range in both groups. 

  + Non-overlapping ranges indicate that there are some regions of confounder space with only cases or only controls in our study. 
  + If we believe the PS model, we can trust the weights and assume this is due to random positivity violations. 
  + In some approaches based on propensity scores, this will cause a problem (matching and subclassification). 
  

---
# Propensity Scores and Positivity

```{r, echo = FALSE, message = FALSE, warning=FALSE, fig.width=10, fig.height=7, fig.align='center'}
dat$ps1 <-predict(fit, type = "response")


ggplot() +
  geom_histogram( aes(x = dat$ps1[dat$qsmk == 0], y = ..density..), fill="#ea8a1a" ) +
  geom_label( aes(x=0.75, y=4, label="Didn't Quit Smoking"),size = 10, label.size = 0.5,  color="#ea8a1a") +
  geom_histogram( aes(x = dat$ps1[dat$qsmk == 1], y = -..density..), fill= "#110566") +
  geom_label( aes(x=0.75, y=-3, label="Quit Smoking"), size = 10, label.size = 0.5, color="#110566") +
  xlim(c(0, 1)) + 
  theme_bw() +
  xlab("Propensity Score") + 
  theme(axis.title = element_text(size = 18))


```



---

## Horvitz-Thompson Estimator

- The *Horvitz-Thompson estimator* is one of two options for estimating $E[Y(a)]$ using our 
estimated weights.

- The Horvitz-Thompson estimator just plugs our estimated weights into our previous formula.

$$
\hat{E}[Y(a)] = \hat{E}\left[\frac{I(A = a)Y}{f(A\vert L)} \right] = \frac{1}{n}\sum_{i = 1}^n I(A_i = a)Y_i W_i
$$

---

## Hajek Estimator

- The *Hajek estimator* fits the linear model 
$$
E[Y \vert A] = \theta_0 + \theta_1 A
$$
by weighted least squares, weighting individuals by our estimated IP weights. 

- It is is equivalent to


$$\frac{ \hat{E}\left[\frac{I(A = a)Y}{f(A\vert L)} \right]}{ \hat{E}\left[\frac{I(A = a)}{f(A\vert L)} \right]} = \frac{\sum_{i = 1}^n I(A_i = a)Y_i W_i}{\sum_{i = 1}^n I(A_i = a) W_i}$$
---

## Hajek Estimator

- Hajek estimator is unbiased for 

$$\frac{ E\left[\frac{I(A = a)Y}{f(A\vert L)} \right]}{ E\left[\frac{I(A = a)}{f(A\vert L)} \right]}$$
- If positivity holds, then  

$$
E\left[\frac{I(A = a)}{f(A\vert L)} \right] = 1
$$

- The Hajek estimator is guaranteed to be between 0 and 1 for dichotomous outcomes. The Horvitz-Thompson estimator is not. 


---

## IP Weighted Effect Estimate in NHEFS

- This code computes the Hajek estimator:

```{r}
f_w_lm <- lm(wt82_71 ~ qsmk, data = dat, weights  = w)
f_w_lm
```

- We estimate that if nobody had quit smoking, the average weight gain would have been 1.8 kg. 

- If everyone had quit smoking, the average weight gain would have been 5.2 kg.

- The average causal effect of quitting smoking on weight gain is 3.4 kg. 

---
## Estimating the Variance of the Estimate

- The raw variance estimate from weighted least squares does not account for uncertainty in the weights. 

- We need to account for having estimated the IP weights. 

- Options: 

  1. Derive the variance analytically
  2. Bootstrap the variance
  3. Use a robust sandwich variance estimate.
  
- Option 3 is conservative but easier than options 1 and 2. 

- Option 3 can be achieved by fitting with GEE using and independent working correlation matrix or using the `HC0` option in `vcovHC`.

---

## Variance Estiamte in the NHEFS Data

```{r, warning=FALSE}
library("geepack")
f_w_gee <- geeglm(wt82_71 ~ qsmk, data = dat,weights = w,
                id = seqn,corstr = "independence")
beta <- coef(f_w_gee)
SE <- coef(summary(f_w_gee))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

```{r}
library(sandwich)
# Equivalent alternative using sandwich
v_wlm_robust <- vcovHC(f_w_lm, type = "HC0")
  
all.equal(v_wlm_robust, vcov(f_w_gee))
```


---
## Stabilized Weights

- Using weights $W^{A} = 1/f(A \vert L)$ we create a pseudo-population in which (heuristically), each person is matched by someone exactly like them who received the opposite treatment. 


- Our pseudo-population is twice as big as our actual sample so the mean of $W^A$ is 2. 

- In the pseudo-population, the frequency of treatment $A = 1$ is 50%. 

- We could have created other pseudo-populations. 

---
## Stabilized Weights

- Our requirements are that, in the pseudo-population, probability of treatment is independent of $L$. 

- But different people could have different probabilities of treatment. 


- To create stabilized weights, we construct a pseudo-population in which the probability of receiving 
each treatment is the same as the frequency of the treatment in our original sample. 

$$
SW^A = \frac{f(A)}{f(A \vert L)}
$$
---

## Stabilized Weights

- In our data, $A$ is binary, so we can compute $f(1)$ as the proportion of quitters in the data. 

```{r}
p1 <- mean(dat$qsmk); p1
```

- In the pseudo-population created by the stabilized weights, each person in our data set corresponds to  26% of a treated person and 74% of an untreated person. 

```{r}
dat <- dat %>% mutate(pa = ifelse(dat$qsmk == 0, 1-p1,  p1),
              sw= pa*w)
summary(dat$sw)
```

- The expected value of $SW^A$ is 1 because the pseudo-population is the same size as
the observed data. 

---

## Estimation Using the Stabilized Weights

```{r}
f_sw_gee <- geeglm(wt82_71 ~ qsmk,data = dat,
  weights = sw, id = seqn,corstr = "independence")
beta <- coef(f_sw_gee)
SE <- coef(summary(f_sw_gee))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

- These are exactly the results we saw with the unstabilized weights. 

---
## Why Use Stabilized Weights

- Differences between stabilized and non-stabilized weights only occur when the model for $E[Y \vert A]$ is not saturated. 


- When the model is not saturated, stabilized weights typically result in greater efficiency. 

---

## Marginal Structural Models

- In the IP weighting strategy we create a population in which $A$ is indepedent of $L$ and then fit the model 
$$E_{ps}[Y \vert A ] = \theta_0 + \theta_1 A$$

- If we believe our conditional exchangeability assumption $Y(a) \ci A \vert L$ and the propensity score model, then in the pseudo-population, $E_{ps}[Y \vert A] = E[Y(a)]$. 

- So the parameter $\theta_1$ is interpretable as the causal risk difference. 

--

- We have proposed a model for the average counterfactual: 

$$E[Y(a)] = \beta_0 + \beta_1 a$$
--

- This model is *marginal* because we have marginalized (averaged) over the values of all
other covariates. 

- It is structural because it is a model for a counterfactual $E[Y(a)]$.

---
## Modeling Continuous Treatments

- With a binary exposure, we could construct a saturated model. 

- For continuous (or highly polytomous) exposures, we can't do that and will have to use 
a parametric model instead. 

- In the NHEFS data, let $A$ be the change in smoking intensity ( $A = 10$ indicates that a person increased their smoking by 10 cigarettes).

- We will limit to only those who smoked 25 or fewer cigarettes per day at baseline ( $N = 1,162$ )

- We can propose a model for our *dose-response curve*

$$
E[Y(a)] = \beta_0  + \beta_1 a + \beta_2 a^2
$$

---
## Estimating Weights for the Continuous Treatment

- To use stabilized weights, we need to estimate $f(A \vert L)$ and $f(A)$. 

- Both of these are PDFs which are hard to estimate!

- We will assume that both $f(A \vert L)$ and $f(A)$ are normal with constant variance across participants.
  + These assumptions are almost certainly wrong. 
  
- We will estimate the mean of $f(A \vert L)$ via linear regression. 

- IP weighted estimates of continuous variables can be very sensitive to choice of weights.

---
## Estimating Weights for the Continuous Treatment

- We fit a linear model for change in smoking intensity including:

  + sex, race, education, exercise, life activity
  + age and age squared 
  + baseline smoking intensity and baseline smoking intensity squared
  + baseline weight and baseline weight squared
  
- The fitted values of this model give us $E[A \vert L]$.

- We can use the variance of the residuals as an estimate of the variance of $A \vert L$

- We then compute $f(A \vert L)$ using `dnorm`. 


---
## Estimating Weights for the Continuous Treatment


```{r}
dat.s <- filter(dat, smokeintensity <= 25)
fal_fit<- lm(smkintensity82_71 ~ as.factor(sex) +
    as.factor(race) + age + I(age ^ 2) +
    as.factor(education) + smokeintensity + I(smokeintensity ^ 2) +
    smokeyrs + I(smokeyrs ^ 2) + as.factor(exercise) + as.factor(active) + wt71 +
    I(wt71 ^ 2), data = dat.s)
fal_mean <- predict(fal_fit, type = "response")
fal  <- dnorm(dat.s$smkintensity82_71, fal_mean,
        summary(fal_fit)$sigma)
```

---
## Estimating Weights for the Continuous Treatment

- To estimate $f(A)$ we use the same procedure, except that the mean of $E[A]$ is is just the average change in smoking intensity. 

```{r}
fa_fit <- lm(smkintensity82_71 ~ 1, data = dat.s)
fa_mean <- predict(fa_fit, type = "response")
fa <- dnorm(dat.s$smkintensity82_71, fa_mean,
            summary(fa_fit)$sigma)
dat.s$sw_cont <- fa / fal
dat.s$w_cont <- 1/fal
```

---
## Estimating Weights for the Continuous Treatment

- For comparison, we compute both the stabilized weights and the non-stabilized weights

```{r}
summary(dat.s$sw_cont)
summary(dat.s$w_cont)
```

- The non-stabilized weights are much more variable!
---
## Fitting the Continuous Treatment Model

- We then fit the marginal structural model $$ E[Y(a)] = \beta_0 + \beta_1 a + \beta_2 a^2$$ using exactly the same strategy we used before. 



```{r, echo = FALSE}
opts <- options(knitr.kable.NA = "")
msm_sw_cont <-geeglm( wt82_71 ~ smkintensity82_71 + I(smkintensity82_71^2),
    data = dat.s, weights = sw_cont, id = seqn, corstr = "independence")
beta <- coef(msm_sw_cont)
SE <- coef(summary(msm_sw_cont))[, 2]
lcl <- beta - qnorm(0.975) * SE 
ucl <- beta + qnorm(0.975) * SE
tabsw <- data.frame(eff = signif(beta, digits = 2) %>% as.character(), 
                    ci = paste0("(",signif(lcl, digits=2), ", ", signif(ucl, digits = 2), ")"), 
                    sp = c(NA, NA, NA))
msm_w_cont <- geeglm( wt82_71 ~ smkintensity82_71 + I(smkintensity82_71^2),
    data = dat.s, weights = w_cont, id = seqn, corstr = "independence")
beta <- coef(msm_w_cont)
SE <- coef(summary(msm_w_cont))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
tabw <- data.frame(eff = signif(beta, digits = 2) %>% as.character(), 
                    ci = paste0("(",signif(lcl, digits=2), ", ", signif(ucl, digits = 2), ")"))
tab <- cbind(tabsw, tabw)
row.names(tab) <- c("Intercept", "Smk Int", "Smk Int Squared")
kable(tab, col.names = c( "Estimate", "95% CI", " ", "Estimate", "95% CI"), 
      format = 'html', row.names = TRUE) %>%
   add_header_above(c(" "=1, "Stabilized Weights" = 2, " " = 1, "Non-Stabilized Weights" = 2))

```
---
## Continuous Model Results


```{r, echo = FALSE,fig.align='center', fig.width=10 }
b <- coef(msm_sw_cont)
fa <- function(a){b[1] + b[2]*a + b[3]*a^2}
ggplot(data.frame(x = range(dat$smkintensity82_71)), aes(x)) + 
   geom_function(fun = fa) + 
  xlab("a") + ylab("E[Y(a)]") + 
  theme(axis.text =  element_text(size = 14), 
        axis.title = element_text(size = 20))
```


---
## Issues with the Continuous Treatment Model

- We had to make strong assumptions about the distribution of $A \vert L$. 

- There are structural positivity issues -- it is not possible for everyone to reduce their smoking by 80 cigarettes per day. 

---
## Effect Modification in Marginal Structural Models

- We might be interested in estimating the causal effect within strata of $V$. 

- We can propose a marginal structural model that includes $V$

$$
E[Y(a) \vert V] = \beta_0 + \beta_1 a + \beta_2 a V + \beta_3 V
$$

- If $V$ and $A$ are both dichotomous, this model is saturated. 

- It is not really fully marginal any more, but we still call it a marginal structural model. 

---
## Interpreting Parameters in the Effect Modification Model

- Suppose that $V$ is sex (0: Male, 1: Female) and $A$ is quitting/not quitting. 

$$
E[Y(a) \vert V] = \beta_0 + \beta_1 a + \beta_2 a V + \beta_3 V
$$

- $\beta_1$ is the causal effect of quitting for men. 

- $\beta_1+ \beta_2$ is the causal effect of quitting for women. 

- What is $\beta_3$?

--

- $\beta_3$ is not a causal parameter because we haven't made any exchangeability assumptions about $V$. 

- $\beta_3$ is the difference between $E[Y(0)]$ in females and $E[Y(0)]$ in males but is not the causal effect of sex. 

---
## Stablilized Weights for the EM Model


- We have two choices for stabilized weights

$$
\frac{f(A)}{f(A \vert L)} \qquad \text{or} \qquad  \frac{f(A\vert V)}{f(A \vert L)}
$$

- Which will be more efficient?

--

- Conditioning on $V$ in the numerator will make the numerator closer to the denominator. 

- The second choice of weights will be less variable. 

- So the second choice of weights is more efficient. 

---
## Stablilized Weights for the EM Model

- We need to compute $f(A \vert V)$ in our data. 

- We can just use the stratified means. 

```{r}
fav_fit <- lm(qsmk ~ sex, data = dat)
dat$pav <- ifelse(dat$qsmk == 0,
                  1-predict(fav_fit, type = "response"), 
                  predict(fav_fit, type = "response"))
dat <- mutate(dat, sw_em = pav*w)
summary(dat$sw_em)
```

---
## Fitting the Effect Modification Model

```{r}
msm_em <- geeglm(
  wt82_71 ~ qsmk*sex, data = dat, weights = sw_em, id = seqn,
  corstr = "independence")
beta <- coef(msm_em)
SE <- coef(summary(msm_em))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

- We have no strong evidence of effect modificaiton by sex. 

---

## Unweighted Regression vs IP Weighting 

- What if we had conditioned on every variable in $L$? 

- The stabilized weights would be 1, so we would just be running a regular 
regression. 

- If we believe in the model for $E[Y(a) \vert L]$, then the regression coefficient
is interpretable as a (conditional) causal parameter. 

- Using the marginal model with IP weighting, we need to believe that our estimates of $f(A \vert L)$ are accurate. 

- In some cases, solving a prediction problem is easier than guessing the correct structural model for all confounders.

---

## Censoring and Missing Data With IP Weights

- In our analysis, we have removed 63 individuals who filled in the survey in 1982 but 
who have a missing value for weight. 


```{r, echo = FALSE}
tab <- dat.u %>% group_by(qsmk) %>% summarize( nmiss = sum(is.na(wt82)), pmiss = round(mean(is.na(wt82))*100, digits = 1))

kable(tab, col.names = c( "Quit Smoking", "Number Missing", "Percent Missing"), 
      format = 'html', row.names = FALSE, digits = 2) 

```

- We have more missing data for people who quit smoking than people who did not, so we could have selection bias. 


---

## Weights for Censoring

- We learned in L4 that, to correct for selection bias, we need to find a set of variables $L_c$ such that $Y(C = 0) \ci C \vert L_c$. 

- We then need to estimate $W^C = 1/P[C = 0 \vert A, L_c]$, or the stabilized version $$SW^C = \frac{P[C = 0 \vert A]}{P[C = 0 \vert A, L_c]}$$

- We will use the same set of variables, we used to compute weights for confounding. 

- The total weights will be $SW^C \cdot SW^A$


---

## Weights for Censoring

- We will estimate $f(C=0 \vert A, L)$ using a logistic model adjusting for
  + the exposure
  + sex, race age, education, exercise, life activity
  + linear and quadratic terms for smoking intensity, duration, and baseline weight. 

- The `dat.u` data frame includes data for all participants. 

```{r}
dat.u$cens <- ifelse(is.na(dat.u$wt82), 1, 0)
fcal_fit <- glm(
  cens ~ as.factor(qsmk) + as.factor(sex) +
    as.factor(race) + age + I(age ^ 2) +
    as.factor(education) + smokeintensity +
    I(smokeintensity ^ 2) + smokeyrs + I(smokeyrs ^ 2) +
    as.factor(exercise) + as.factor(active) + wt71 + I(wt71 ^ 2),
  family = binomial(),
  data = dat.u
)
fc0al <- 1 - predict(fcal_fit, type = "response")
```

---

## Weights for Censoring

- We can estimate the numerator of the stablilized weight, $f(C = 0 \vert A)$ using the average within levels of $A$. 

```{r}
fca_fit <-glm(cens ~ as.factor(qsmk), family = binomial(), data = dat.u)
fc0a <- 1 - predict(fca_fit, type = "response")
sw.c <- fc0a/fc0al
dat$sw.c <- sw.c[!is.na(dat.u$wt82)]
dat$sw.comb <- dat$sw*dat$sw.c
```


```{r}
summary(dat$sw.c)
summary(dat$sw.comb)
```

---
## Combined Weights Results

- Below are results using the combined weights for both censoring and confounding

```{r, echo = FALSE}
msm.sw <- geeglm(wt82_71 ~ qsmk, data = dat,weights = sw.comb,
  id = seqn,  corstr = "independence")
beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
tab <- data.frame(eff = signif(beta, digits = 3) %>% as.character(), 
                    ci = paste0("(",signif(lcl, digits=3), ", ", signif(ucl, digits = 3), ")"))
row.names(tab) <- c("Intercept", "Quit Smoking")
kable(tab, col.names = c( "Estimate", "95% CI"), 
      format = 'html', row.names = TRUE) 
```

- Compared with the previous results

```{r, echo = FALSE}
msm.sw <- geeglm(wt82_71 ~ qsmk, data = dat,weights = sw,
  id = seqn,  corstr = "independence")
beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
tab <- data.frame(eff = signif(beta, digits = 3) %>% as.character(), 
                    ci = paste0("(",signif(lcl, digits=3), ", ", signif(ucl, digits = 3), ")"))
row.names(tab) <- c("Intercept", "Quit Smoking")
kable(tab, col.names = c( "Estimate", "95% CI"), 
      format = 'html', row.names = TRUE) 
```

---
## Survey Weights

- Typically, nationally representative surveys won't use a completely random sampling scheme. 
  + They will try to over-sample smaller groups. 
  + This allows them to get better estimates for less common covariate values. 
  
- Over-sampling introduces selection bias that needs to be corrected.

- If the survey was done well, a specific over-sampling rule was used. 
  + For example, you might select census tracts randomly according to a covariate informed probability and then select housees randomly within census tract. 
  
- This allows the survey administers to compute $P[C = 0 \vert L]$ for every participant. 


---
## Survey Weights


- If $L$ includes every factor affecting study inclusion, then $P[C = 0 \vert A, L] = P[C = 0 \vert L]$ for any exposure $A$. 

- So we could use the same set of weights regardless of what exposure we are looking at. 

- Usually, survey administers will compute $P[ C = 0 \vert L]$ and distribute $1/P[C = 0 \vert L]$ as "survey weights". 
  + Survey weights are censoring weights!

- If you have survey data that come with survey weights, be sure to read about how these are calculated. 
  + Most likely, if you want to estimate a parameter for the larger population, you will need to use the survey weights. 


---
# 3. G-Computation (Standardization)
---

## Standardization

- In standardization, we compute an estimate of $E[Y \vert A = a, L = l]$ and plug it into the g-formula:

$$
E[Y(a)] = \sum_l E[Y \vert A = a, L = l]P[L = l]
$$

- If $L$ is continuous, we need to integrate

$$
E[Y(a)] = \int_l P[Y \vert A = a, L = l]f(l) dl
$$

---

## Outcome Modeling

- In the standardization formula, we need to estimate $P[Y \vert A = a, L = l]$.

- To do this we fit a linear regression

```{r}
f_y <-glm( wt82_71 ~ qsmk + sex + race + age + I(age * age) + as.factor(education)
    + smokeintensity + I(smokeintensity * smokeintensity) + smokeyrs
    + I(smokeyrs * smokeyrs) + as.factor(exercise) + as.factor(active)
    + wt71 + I(wt71 * wt71) + qsmk * smokeintensity,
    data = dat.u)
summary(fit)$coefficients %>% head()
```
---
## Outcome Modeling

- This first step of fitting the outcome model is just fitting a regular linear regression. 

- If we believe the model, then the coefficient for treatment gives a *conditional* causal effect estimate within strata of $L$. 

- If there were no interactions between $A$ and $L$ in the model, the coefficient on $A$ would give the marginal treatment effect. 

  + This is true using a linear model but would not be true in a model with a non-linear link.


- An outcome model with no interactions assumes **homogeneity**, i.e. the average treatment effect is the same in every level of $L$. 

  + Typically, homogeneity is considered to be a very strong assumption.

---

## Outcome Modeling


- Our outcome model included an interaction between $A$ and smoking intensity.

- In this model, the coefficient on $A$ is the average treatment effect within strata of smoking intensity. 

- To estimate the average (marginal) treatment effect, we need to average over the distribution of smoking intensity, or more generally the distribution of $L$. 


---
## Standardizing (G-Computation)

- Fortunately, we do not need to compute $f(L)$.


- Using the definition of expectation, we can write
$$
 \int_l P[Y \vert A = a, L = l]f(l)dl = E\left[ E\left[Y \vert A = a, L = l \right]\right]
$$
- So we can estimate the standardized mean as

$$
\frac{1}{n}\sum_{i=1}^n\hat{E}[Y \vert A = a, L_i]
$$
---

## Standardizing

- For each person, we need to compute two values $\hat{E}[Y \vert A = 0, L_i]$ and $\hat{E}[Y\vert A = 1, L_i]$

- These are predicted values that we can get out of our previous regression. 

- We predict each person's outcome with $A = 0$ and with $A = 1$. 

```{r}
# Make two copies of the data
dat0 <- dat1 <- dat.u 
# In one copy everyone got treatment 0
dat0$qsmk <- 0
# In the second copy everyone got treatment 1
dat1$qsmk <- 1

Y0 <- predict(f_y, newdata = dat0, type = "response") %>% mean()
Y1 <- predict(f_y, newdata = dat1, type = "response") %>% mean()
cat(round(Y1, digits = 2), "-", round(Y0, digits = 2), "=", round(Y1-Y0, digits = 2))
```

---
## Censoring

- If the set of variables in our model for $E[Y \vert A, L]$ are sufficient to adjust for censoring (meaning that they block all paths between $Y(C = 0)$ and $C$), then we don't need to do anything extra to adjust for censoring.

- The trick is to predict $Y(a)$ for **all individuals** with covariate data, not just those with non-missing outcomes. 

- If $L$ is not sufficient to adjust for censoring, we may need to add additional variables specifically to deal with censoring. 

- Recall that the s-backdoor criterion gives us a way to select variables to condition on. 

---

## Estimating the Variance 

- To compute the variance of the estimate, we need to bootstrap. 

- We will repeatedly re-sample our data with replacement. 

- Each time, we re-fit the outcome model and recompute the standardized effect. 

- The standard deviation of our bootstrap samples consistently estimates the standard error of our estimate. 

---
## Bootstrapping the Variance 

- Step 1: Write a function to compute the standardized mean
```{r}
# Step 1: Write a function compute the standardized mean
std_mean_func <- function(d){
   boot_fit <-glm( wt82_71 ~ qsmk + sex + race + age + I(age * age) + as.factor(education)
    + smokeintensity + I(smokeintensity * smokeintensity) + smokeyrs
    + I(smokeyrs * smokeyrs) + as.factor(exercise) + as.factor(active)
    + wt71 + I(wt71 * wt71) + qsmk * smokeintensity, data = d)
  d0 <- d1 <- d
  d0$qsmk <- 0 
  d1$qsmk <- 1
  Y0 <- predict(boot_fit, newdata = d0, type = "response") %>% mean()
  Y1 <- predict(boot_fit, newdata = d1, type = "response") %>% mean()
  return(Y1-Y0)
}
```


---
## Bootstrapping the Variance 

- Step 2: Repeatedly re-sample the data and compute the standardized mean in the re-sampled data set. 

```{r}
set.seed(1)
samples <- replicate(n = 500,
                     expr = sample(seq(nrow(dat.u)), 
                                   size = nrow(dat.u), replace = TRUE))
res <- apply(samples, 2, FUN = function(ix){
  boot_dat <- dat.u[ix,]
  boot_est <- std_mean_func(boot_dat)
  return(boot_est)
})
se <- sd(res)
eff <- Y1-Y0
ci <- eff + c(-1, 1)*qnorm(0.975)*se 
cat(round(eff, digits = 2), "(", round(ci, digits = 2), ")")
```



---
## Effect Modification in G-Computation

- If we want to estimate the average causal effect stratified by $V$, we can simply stratify by $V$ before averaging our predicted values. 

```{r}
dat.u$Y0 <- predict(fit, newdata = dat0, type = "response") 
dat.u$Y1 <- predict(fit, newdata = dat1, type = "response")
dat.u %>% group_by(sex) %>% 
      summarize(Y0 = mean(Y0), Y1 = mean(Y1))
```


---

## IP Weighting vs G-Computation

- If all of our models are fully saturated, IP weighting and standardization give exactly the same result (recall HW1). 

- With parametric models, the two methods rely on different assumptions and are not the same.

- In the IP weighted analysis, we need to trust our model for $f(A \vert L)$.

- In the standardized analysis, we need to trust our model for $E[Y \vert A, L]$. 

- In either case, if our our parametric models are wrong, our predictions will be biased and so our estimators will be biased. 

- May be wise to compute both and compare.


---

## IP Weighting vs G-Computation

- Chatton et al (2020) present simulation results comparing IPW, G-computation, and two methods we haven't talked about yet. 

- In their simulation scenario, G-computation always had a lower MSE than IPW 

- For IPW, they calculated variance using robust SEs (as we did earlier) and found slight over coverage (~97% for a 95% target). 

- They used parametric simulation to compute the variance of the G-computation estimate. This gave close to target coverage but slightly under in some cases (~94%). 


---

## Chatton et al Results



```{r, echo = FALSE, message = FALSE, warning=FALSE, fig.width=10, fig.height=4, fig.align='center'}
library(gridExtra)
chatton_data <- data.frame(N = rep(c(100, 300, 500, 2000 ), each = 8), method = rep(rep(c("GC", "IPW"), each = 4), 4), 
                           covars = rep(c("outcome", "treatment", "common", "all"), 16), 
                           mse = c(0.526, 0.580, 0.552, 0.558, 
                                   0.578, 0.716, 0.587, 0.741, 
                                   0.283, 0.319, 0.304, 0.297, 
                                   0.301, 0.372, 0.318, 0.361,
                                   0.217, 0.245, 0.233, 0.228,
                                   0.230, 0.285, 0.244, 0.274, 
                                   0.108, 0.122, 0.116, 0.113, 
                                   0.113, 0.138, 0.120, 0.131), 
                           power = c(17.7, 14.0, 15.1, 16.9, 
                                     7.8, 9.8, 8.8, 9.6, 
                                     43.6, 35.2, 38.5, 39.9, 
                                     24, 21.4, 26.1, 20.0, 
                                     64.5, 53.6,57.6,  60.5,
                                     43.3, 35.4, 43.7, 33.9, 
                                     99.6, 98.6, 99.1, 99.4, 
                                     98.1, 93.0, 97.7, 93.6))
p1 <- chatton_data %>% filter(covars == "common") %>% 
ggplot() + geom_point(aes(x = N, y = mse, shape = method, color = method), size = 3) + 
  geom_line(aes(x = N, y = mse, color = method)) + 
  ylab("MSE") + 
  theme_bw() + 
  theme() + 
   theme(axis.title = element_text(size = 18), 
         legend.text = element_text(size  = 18), 
         legend.title = element_text(size = 18), 
         axis.text = element_text(size = 14))

p2 <- chatton_data %>% filter(covars == "common") %>% 
ggplot() + geom_point(aes(x = N, y = power, shape = method, color = method), size = 3) + 
  geom_line(aes(x = N, y = power, color = method)) + 
  ylab("Power") + 
  theme_bw() + 
  theme() + 
   theme(axis.title = element_text(size = 18), 
         legend.text = element_text(size  = 18), 
         legend.title = element_text(size = 18), 
         axis.text = element_text(size = 14))
grid.arrange(p1, p2, ncol = 2)
```


---

## IP Weighting vs G-Computation

- Snowden et al (2010) describe G-Computation as rarely used in epidemiology due to lack of awareness. 

- However, G-Computation can be more precise than IP weighting in some circumstances. 

  + IP weighting can be inefficient when weights are very large. 
  
  
- Vansteelandt and Keiding find that IPW performs better when the covariate distribution is very different in cases and controls. 
  + Depending on the model used fo G-Computation, the dominant group may have too much influence over estimates in these regions. 
  + G-Computation may under-estimate the uncertainty about the counterfactuals in these regions. 
  
  
- Some  suggest that we shouldn't be trying to estimate the countefactuals in these regions at all, since this requires extrapolating into regions with low data density. 
