---
title: "In Class Acitivites"
date: "Last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

$\newcommand{\ci}{\perp\!\!\!\perp}$
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_search(show_icon = TRUE)
xaringanExtra::use_panelset()
```

```{r xaringanthemer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  link_color = "#ea8a1a",
  base_color = "#110566",
 # header_font_google = google_font("Josefin Sans"),
 # text_font_google   = google_font("Montserrat", "300", "300i"),
 # code_font_google   = google_font("Fira Mono")
)
```


## Monday, February 5, 2025

- We are interested in the effect of a genetic variant, $G$ on an outcome $Y$. 

- Participants who require surgery, either for $Y$ or a different condition are recruited into a genetic study. 

- The variant $G$ could affect $Y$ and could also affect other medical conditions that require surgery. 

- Draw a DAG of this scenario.

- Would you expect selection bias to affect the estimate of $G$ on $Y$? Why/why not?


---
## Monday, February 19

Match the estimator with statements that are true about it. 
.pull-left[
**Estimator**

- Hajek Estimator

- G-Computation

- AIPW estimator from Robins, Rotnitzky and Zhao (1994)

- Subclassification estimator with fixed $K$

- Bang and Robins estimator

- Adjusting for propensity score as a linear covariate. 
]

.pull-right[

** Statements**

- Consistent if PS model is correct

- Consistent if outcome model is correct

- Not consistent 

- These two are asymptotically equivalent

- More accurate using estimated PS than using true PS.
]

---
## Wednesday, February 26, 2025

Multiple choice questions to do together.

- When you first begin thinking about a causal question you should:

  A. Draw a DAG containing all confounders of the exposure and outcome that you can think of, along with any sources of censoring that might affect your study. 
  
  B. Draw a DAG that only contains variables that are measured and in your data.
  
  C. Draw a DAG containing both variables that are measured and variables that you think that you could measure.

---
## Wednesday, February 26, 2025


- When looking at your DAG, you find that there are confounders that you cannot measure. You should:

  A. Remove these variables from the DAG, go ahead with your analysis as though they did not exist and report your estimated causal effect. 
  
  B. Go ahead with your analysis but do a sensitivity analysis to understand how much bias your unmeasured confounders could be causing.
  
  C. Look for an alternative way to measure the causal effect like instrumental variable analysis. 
  
  D. Give up on measuring a causal effect and do a descriptive analysis describing associations.


---
## Wednesday, (Date)


Review of IVA concepts:

In pairs, write down a definition for one of these terms and how it relates to IVA:

- Relevance

- Exclusion Restriction

- Local average treatment effect

- Weak instrument bias

- Two Stage Least Squares

- IVW Regression

