---
title: "In Class Acitivites"
date: "Last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

$\newcommand{\ci}{\perp\!\!\!\perp}$
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_search(show_icon = TRUE)
xaringanExtra::use_panelset()
```

```{r xaringanthemer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  link_color = "#ea8a1a",
  base_color = "#110566",
 # header_font_google = google_font("Josefin Sans"),
 # text_font_google   = google_font("Montserrat", "300", "300i"),
 # code_font_google   = google_font("Fira Mono")
)
```


## Monday, February 5, 2024

- We are interested in the effect of a genetic variant, $G$ on an outcome $Y$. 

- Participants who require surgery, either for $Y$ or a different condition are recruited into a genetic study. 

- The variant $G$ could affect $Y$ and could also affect other medical conditions that require surgery. 

- Draw a DAG of this scenario.

- Would you expect selection bias to affect the estimate of $G$ on $Y$? Why/why not?


---
## Monday, (Date)

Match the estimator with statements that are true about it. 
.pull-left[
**Estimator**

- Hajek Estimator

- G-Computation

- AIPW estimator from Robins, Rotnitzky and Zhao (1994)

- Subclassification estimator

- Bang and Robins estimator

- Adjusting for propensity score as a linear covariate. 
]

.pull-right[

** Statements**

- Consistent if PS model is correct

- Consistent if outcome model is correct

- Not consistent 

- These two are asymptotically equivalent

- More accurate using estimated PS than using true PS.
]


---
## Wednesday, (Date)


Review of IVA concepts:

In pairs, write down a definition for one of these terms and how it relates to IVA:

- Relevance

- Exclusion Restriction

- Local average treatment effect

- Weak instrument bias

- Two Stage Least Squares

- IVW Regression

